Distortion-corrected kernel density estimate on Riemannian manifolds

### 2022-09-26 paper/paper.tex
Remove C and colors for 4 kernels

As pointed out by \textcite{Perrault-Joncas2013-pq}, the naive distance \(\|\pmb{p}-\pmb{q}\|\) underestimates \(d_g\), and the graph shortest distance \(d_G\) overestimates \(d_g\) by the triangle inequality. By using only the first-order correction in \(\hat{d}_g\), our estimator is also affected by the estimated Riemannian metric. In order to correct for the estimation error in \(\pmb{H}(\pmb{y}_i)\), we further divide the density estimator in \eqref{eq:denestimator} by a constant term \(C\) so that the integral of the kernel is equal to 1. The constant term in \(1\)-d is given by the \emph{area under the curve (AUC)}, and can be named as the \emph{volume between the surface (VBS)} in higher dimension. We use the volume estimator in Section 6.4 of \textcite{Perrault-Joncas2013-pq} to sum over all embedding points \(\pmb{y}_i\) and the estimated point \(\pmb{p}\) so that for \(\pmb{Z} = \{\pmb{y}_1, \dots, \pmb{y}_N, \pmb{p}\}\), we have
\begin{equation}
\label{eq:vbs}
C(\pmb{p}) = \sum_{i=1}^{N+1} \sqrt{\det(\pmb{H}(\pmb{z}_i))} \Delta z^1(\pmb{z}_i)\dots \Delta z^d(\pmb{z}_i),
\end{equation}
where \(\pmb{z}_i=(z^1(\pmb{z}_i),\dots, z^d(\pmb{z}_i))\). The distortions are corrected for the tangent plane of the embedding points. Since we aim to estimate the density of the manifold for the embedding of sample points, \(\pmb{p}\) is equal to \(\pmb{y}_i\), for \(i=1,\dots,N\). Therefore, we have
\begin{equation}
\label{eq:vbsyi}
C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i).
\end{equation}

@@

We first generate a 2-dimensional data of size \(N=2000\) from a Gaussian mixture model with four components with different means
\(\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime\) and the same variance-covariance matrix \(\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4\). The mixture proportions
are equally set as \(\pi_i=0.25, i=1,2,3,4\).
Then the mixture Gaussian mixture density function is a weighted linear combination of the four component Gaussian densities as
\begin{equation}
\label{eq:gmm}
P(\pmb{X}=\pmb{x}) = \sum_{i=1}^{4}\pi_i \frac{1}{(2\pi)^{d/2}|\pmb{\Sigma}_i|^{-1/2}} \exp{\{-\frac{1}{2} (\pmb{x}-\pmb{\mu_i})^\prime \pmb{\Sigma}_i^{-1} (\pmb{x}-\pmb{\mu_i}) \}}.
\end{equation}

\caption{True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.02, 0.02)$. The colors indicate the density of the data and lower density points in darker colors are scattered both in the outer and center areas. The shapes indicate the four kernels.}\label{fig:metadensity}
\end{figure}

@@

@@

{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_isomap_r0_5_rank}

@@

\caption{Scatterplot of true density and estimated density ranks of ISOMAP embedding for KDE with both variable and fixed bandwidth. The four colors and shapes represents the four gaussian kernels in the 2-D meta data. DC-KDE shows a strong linear positive relationship.}\label{fig:tpisomapden}
\end{figure}

>>>>

@@

We first generate a 2-dimensional data of size \(N=2000\) from a Gaussian mixture model with four components with different means
\(\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime\) and the same variance-covariance matrix \(\pmb{\Sigma}_i=diag(0.016, 0.016), i=1,2,3,4\). The mixture proportions
are equally set as \(\pi_i=0.25, i=1,2,3,4\).
Then the mixture Gaussian mixture density function is a weighted linear combination of the four component Gaussian densities as
\begin{equation}
\label{eq:gmm}
P(\pmb{X}=\pmb{x}) = \sum_{i=1}^{4}\pi_i \frac{1}{(2\pi)^{d/2}|\pmb{\Sigma}_i|^{-1/2}} \exp{\{-\frac{1}{2} (\pmb{x}-\pmb{\mu_i})^\prime \pmb{\Sigma}_i^{-1} (\pmb{x}-\pmb{\mu_i}) \}}.
\end{equation}

@@

\caption{True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.016, 0.016)$. The colors indicate the density of the data and lower density points in darker colors are scattered both in the outer and center areas. The shapes indicate the four kernels.}\label{fig:metadensity}
\end{figure}

We use the multivariate KDE with a fixed plug-in bandwidth in \autoref{mvkde} as the optimal density estimates of the manifold to test the proposed DC-KDE. To compare both density estimation methods, it is essential to know the \emph{true density} of the 2-D manifold. The Gaussian mixture distribution is known from \eqref{eq:gmm}, however, it is not the density of the manifold because of the mapping function in \autoref{eq:twinpeak} distorts the distribution and the manifold of the density is no longer the Gaussian mixture density. Therefore, we apply DC-KDE to the 3-D twin peaks data and set the embedding as the 2-D meta data (true manifold) to first calculate the true density of the manifold. We use this technique to obtain the true density of the manifold throughout the simulation process.

@@

{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_isomap_r0_5_rank_nocolor}

@@

\caption{Scatterplot of true density and estimated density ranks of ISOMAP embedding for KDE with both variable and fixed bandwidth. DC-KDE shows a strong linear positive relationship.}\label{fig:tpisomapden}
\end{figure}
